Demystifying the Computer Science Ph.D. Admission in the US \\{\Large A guide for Vietnamese and International Students

### 2023-06-11 phd-cs-us.tex
updates

As mentioned, if you're admitted to a \emph{good} CS PhD program, you should not have to worry about funding!  
In the US, the common types of funding for PhD are \emph{graduate teaching assistant} (GTA or TA), \emph{graduate research assistant} (GRA or RA), and \emph{Fellowship}.
RA is the type of funding paid by a prof. for you to do their research. TA is paid by the department for you to help with teaching. Finally, fellowship is an independent funding that can come the school, a company, or organization. Tab.~\ref{tab:funding} summarizes the differences.

@@

\paragraph{How to get TA?}  Unless you have other funding such as RA or Fellowships, TA is typically a default thing. When you apply to be a full-time student, you also sate that you need financial assistant. It is common that the PhD committee will either admit you and give you GTA, or reject you; i.e., we do not admit a student without supporting them.

@@

\begin{enumerate}   
    \item What can you do to increase your admission chance?
    \begin{description}
        \item Show something that makes you \textbf{stand out}, e.g., are you a female or a minority in CS (research for this on Google)? Do you participate in outreach activities that help increase diversity and inclusion in CS?  All of these are important in CS in the US.

    \item Even if you do not have formal research experience, you can talk about your personal projects. For example, if you have an open-source project on Github that is used by many people, has lots of stars in Github, do talk about it. If you write technical, research-like blogs, talk about them too.
    \end{description}

@@

\begin{description}
\item International students not familiar with US universities often put them into \emph{two} bins:  (i) very top school that they dream about such as Stanford, MIT, Princeton, Harvard and (ii) everything else.  Sometimes they apply reputations of schools for non-CS programs, e.g., medical or physics, to rank CS programs. 
In some cases they rank universities based on popular states they know in the US, e.g., California, New York, and Texas.  Let's just say there are so many thing wrong with these methods.

@@

\item But more generally, rankings is superficial and you will want to do more research to be informed and make better decision. For example, if you get admissions to several places, you should contact profs. that you're interested in at those place and talk to them. They would be more willing to chat to you now that you have been admitted.  Ask them questions about their work, how they manage students, their expectations. You can even ask to contact their students.
\end{description}

    \item Can I apply to PhD to CS if my undergrad was not in CS or any related areas?
    \begin{description}
        \item Yes, as long as you can demonstrate you are ready for CS PhD research through research experiences, LoRs, statements, etc as mentioned. You might be even able leverage this to make your profile stand out.
    \end{description}

    \item   Is an MS degree required for admission to PhD?
    \begin{description}
    \item No. In fact, student with BS can get MS degree "along the way" to PhD.  However, MS can help if it gives research experience or is from a more well-known school than your undergrad institution.
    \end{description}

    \item How do I address a professor Firstname Lastname?
    \begin{description}
        \item If you don't know that professor (e.g., first email contact), then use \textbf{Prof. Lastname} or \textbf{Dr. Lastname}. I've seen many international students write \textbf{Prof.} or \textbf{Dr.} \textbf{Firstname Lastname}.  Writing like that makes it like you copy and paste the names, so no need to do so,  just Prof. or Dr. Lastname.

@@

\begin{commentbox}
    Vu: I've been called Dr. Vu and I find it a bit amusing but am totally fine with it.
    \end{commentbox}
    \end{description}
\end{enumerate}

>>>>

As mentioned, if you're admitted to a \emph{good} CS PhD program, you should not have to worry about funding!  
In the US, the common types of funding for PhD are \emph{graduate teaching assistant} (GTA or TA), \emph{graduate research assistant} (GRA or RA), and \emph{Fellowship}.
RA is the type of funding paid by a prof. for you to do their research. TA is paid by the department for you to help with teaching. Finally, fellowship is an independent funding that can come from a school, a company, or an organization. Tab.~\ref{tab:funding} summarizes the differences.

@@

\paragraph{How to get TA?}  Unless you have other funding such as RA or Fellowships, TA is typically a default thing. When you apply to be a full-time student,  state that you need financial assistant. It is common that the PhD committee will either admit you and give you GTA, or reject you; i.e., we do not admit a student without supporting them.

@@

\begin{enumerate}   
    \item What can you do to increase your admission chance?
      \begin{itemize}[leftmargin=*]
        \item Show something that makes you \textbf{stand out}, e.g., are you a female or a minority in CS (research for this on Google)? Do you participate in outreach activities that help increase diversity and inclusion in CS?  All of these are important in CS in the US.

    \item Even if you do not have formal research experience, you can talk about your personal projects. For example, if you have an open-source project on Github that is used by many people, has lots of stars in Github, do talk about it. If you write technical, research-like blogs, talk about them too.
    \end{itemize}

@@

\begin{itemize}[leftmargin=*]
\item International students not familiar with US universities often put them into \emph{two} bins:  (i) very top school that they dream about such as Stanford, MIT, Princeton, Harvard and (ii) everything else.  Sometimes they apply reputations of schools for non-CS programs, e.g., medical or physics, to rank CS programs. 
In some cases they rank universities based on popular states they know in the US, e.g., California, New York, and Texas.  Let's just say there are so many thing wrong with these methods.

@@

\item But more generally, rankings is superficial and you will want to do more research to be informed and make better decision. For example, if you get admissions to several places, you should contact profs. that you're interested in at those place and talk to them. They would be more willing to chat to you now that you have been admitted.  Ask them questions about their work, how they manage students, their expectations. You can even ask to contact their students.
\end{itemize}

    \item Can I apply to PhD to CS if my undergrad was not in CS or any related areas?
    \begin{itemize}[leftmargin=*]
        \item Yes, as long as you can demonstrate you are ready for CS PhD research through research experiences, LoRs, statements, etc as mentioned. You might be even able leverage this to make your profile stand out.
    \end{itemize}

    \item   Is an MS degree required for admission to PhD?
    \begin{itemize}[leftmargin=*]
    \item No. In fact, student with BS can get MS degree "along the way" to PhD.  However, MS can help if it gives research experience or is from a more well-known school than your undergrad institution.
    \end{itemize}

    \item How do I address a professor Firstname Lastname?
    \begin{itemize}[leftmargin=*]
        \item If you don't know that professor (e.g., first email contact), then use \textbf{Prof. Lastname} or \textbf{Dr. Lastname}. I've seen many international students write \textbf{Prof.} or \textbf{Dr.} \textbf{Firstname Lastname}.  Writing like that makes it like you copy and paste the names, so no need to do so,  just Prof. or Dr. Lastname.

@@

\begin{commentbox}
    Vu: I've been called Dr. Vu and I find it a bit amusing but am totally fine with it.
    \end{commentbox}
    \end{itemize}
\end{enumerate}
